Computing Exact Restricted Isometry Constants via Mixed Integer Semidefinite Programming

### 2016-03-18 RIP-MISDP-Paper/IEEE_RIP-MISDP.tex
paper: reformulation for aborts
mixed-integer -> mixed integer

\usepackage[breaklinks,colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue,bookmarks=false,hypertexnames=false,
  pdftitle={Computing Exact Restricted Isometry Constants via Mixed Integer Semidefinite Programming},
  pdfauthor={T. Gally, and M. E. Pfetsch}]{hyperref}

@@

\title{Computing Exact Restricted Isometry Constants via Mixed Integer Semidefinite Programming}
%\title{Computing Restricted Isometry Constants Exactly via Mixed Integer Semidefinite Programming}
\author{Tristan~Gally and Marc~E.~Pfetsch
  \thanks{This work was supported by the German Research Foundation (DFG) through Collaborative Research Center 805 and Priority Program 1798.}
  \thanks{The  authors  are  with  the  Research  Group  Optimization,  TU  Darmstadt,
  Darmstadt  64293,  Germany  (e-mail:  gally@mathematik.tu-darmstadt.de;
  pfetsch@mathematik.tu-darmstadt.de).}}

@@

\noindent
\begin{abstract}
  One of the fundamental tasks in compressed sensing is finding the
  sparsest solution to an underdetermined system of linear equations. It is
  well known that although this problem is \NP-hard, under certain
  conditions it can be solved by using a linear program which minimizes the
  1-norm. The restricted isometry property has been one of the key
  conditions in this context. However, computing the best constants for
  this condition is itself \NP-hard. In this paper we propose a
  mixed-integer semidefinite programming approach for computing these
  optimal constants. This also subsumes sparse principal component
  analysis. Computational results with this approach allow to evaluate
  earlier semidefinite relaxations and show that their quality is limited.
\end{abstract}

@@

The goal of this paper is to compute optimal restricted isometry constants
via mixed-integer semidefinite programming (MISDP). To this end, we will
introduce an MISDP with a rank-1 constraint. In order to utilize this
MISDP, we will show that the rank constraint can be relaxed without
changing the optimal value. We can then apply an MISDP solver.

@@

Theorem~\ref{Rk1thm} guarantees that \eqref{MISDP} always has a solution of
rank one. Therefore, the optimal values of \eqref{MISDP} and \eqref{QP}
agree. We can therefore compute the optimal constant $\alpha_k$ for the RIP
by solving \eqref{MISDP}. Since the same argumentation also holds for the
maximization problem for $\beta_k$, this allows us to compute the
restricted isometry constant and the restricted isometry ratio by
mixed-integer semidefinite programming.

@@

\noindent
In this section we demonstrate the applicability of the proposed MISDP-formulation and use it to
assess the quality of the relaxations proposed in~\cite{Asp08} and~\cite{Asp07}.
For solving the mixed-integer semidefinite programs we use the MISDP solver SCIP-SDP~\cite{SCIP-SDP},
originally developed in~\cite{Mar13} and~\cite{MS12}. It combines the
branch and bound framework SCIP 3.2.1 \cite{SCIP} with 
interior-point SDP-solvers, in our case SDPA~7.3.8~\cite{SDPA6,SDPA7}.

@@

The tests were performed on a Linux cluster with Intel i3 CPUs with 3.2GHz, 4MB cache, and 8GB memory running
Linux. Each computation was performed single-threaded with a single process running on each computer and with 
a time limit of four hours. The code was compiled with gcc 4.4.5 with \texttt{-O3} optimization. For 
Problem~\eqref{Asp08}, the 585 evaluations of~\eqref{Asp08Dual} for 39 matrices and all 15 choices of $\rho$
were aborted because of the memory limit of 8GB. These are counted with the maximum allowed time of four hours.

@@

The solving times in seconds for the left- and right-hand side of the RIP are given in
Table~\ref{lhsRhsTime}. The relaxation~\eqref{Asp07}
is very fast, taking only a few seconds for all instances. The mixed-integer semidefinite program obviously
cannot compete with these solving times, although for band-matrices, the difference is relatively
small. The Lagrangian Relaxation~\eqref{Asp08}, however, takes more time to produce bounds than is needed
by~\eqref{MISDP} to compute exact solutions. This is mainly caused by the fact that~\eqref{Asp08Dual} involves
$n+1$ matrix variables of dimension $n\times n$ with corresponding SDP-constraints instead of a 
single one each in~\eqref{MISDP} and~\eqref{Asp07}.
Furthermore, the big difference is also caused by the instances running into the 
memory limit, but even without those, the algorithm still takes slightly longer than~\eqref{MISDP}.

@@

\noindent
This paper shows that exact values of restricted isometry constants can be
computed via mixed-integer semidefinite programming. The computed bounds
significantly improve the bounds computed by the semidefinite
relaxations~\eqref{Asp07} and~\eqref{Asp08}
proposed in~\cite{Asp07} and \cite{Asp08}, respectively. In conclusion, the relaxation~
\eqref{Asp07} is fast to compute, but produces a gap of around 90\,\% for
$\alpha_k$ and 30\,\% for $\beta_k$. If successfully solved, 
\eqref{Asp08} provides better bounds for $\beta_k$ for most types of 
matrices, but its running time is too large
to be used in practice. Moreover, directly solving~\eqref{MISDP} is often
faster and produces the exact value. Consequently, 
\eqref{Asp08} should not be used in practice.

Still, the matrices currently handable by~\eqref{MISDP} are small. For the
future, the hope is that with the advancement of solving techniques for
mixed-integer semidefinite programs, the sizes of the matrices can be
significantly increased. The components to be improved include cutting
planes, primal heuristics, and branching rules.

>>>>

\usepackage[breaklinks,colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue,bookmarks=false,hypertexnames=false,
  pdftitle={Computing Exact Restricted Isometry Constants via Mixed Integer Semidefinite Programming},
  % pdftitle={Computing Restricted Isometry Constants via Mixed Integer Semidefinite Programming},
  pdfauthor={T. Gally, and M. E. Pfetsch}]{hyperref}

@@

\title{Computing Exact Restricted Isometry Constants via Mixed Integer Semidefinite Programming}
%\title{Computing Restricted Isometry Constants via Mixed Integer Semidefinite Programming}
\author{Tristan~Gally and Marc~E.~Pfetsch
  \thanks{This work was supported by the German Research Foundation (DFG) through Collaborative Research Center 805 and Priority Program 1798.}
  \thanks{The  authors  are  with  the  Research  Group  Optimization,  TU  Darmstadt,
  Darmstadt  64293,  Germany  (e-mail:  gally@mathematik.tu-darmstadt.de;
  pfetsch@mathematik.tu-darmstadt.de).}}

@@

\noindent
\begin{abstract}
  One of the fundamental tasks in compressed sensing is finding the
  sparsest solution to an underdetermined system of linear equations. It is
  well known that although this problem is \NP-hard, under certain
  conditions it can be solved by using a linear program which minimizes the
  1-norm. The restricted isometry property has been one of the key
  conditions in this context. However, computing the best constants for
  this condition is itself \NP-hard. In this paper we propose a
  mixed integer semidefinite programming approach for computing these
  optimal constants. This also subsumes sparse principal component
  analysis. Computational results with this approach allow to evaluate
  earlier semidefinite relaxations and show that their quality is limited.
\end{abstract}

@@

The goal of this paper is to compute optimal restricted isometry constants
via mixed integer semidefinite programming (MISDP). To this end, we will
introduce an MISDP with a rank-1 constraint. In order to utilize this
MISDP, we will show that the rank constraint can be relaxed without
changing the optimal value. We can then apply an MISDP solver.

@@

Theorem~\ref{Rk1thm} guarantees that \eqref{MISDP} always has a solution of
rank one. Therefore, the optimal values of \eqref{MISDP} and \eqref{QP}
agree. We can therefore compute the optimal constant $\alpha_k$ for the RIP
by solving \eqref{MISDP}. Since the same argumentation also holds for the
maximization problem for $\beta_k$, this allows us to compute the
restricted isometry constant and the restricted isometry ratio by
mixed integer semidefinite programming.

@@

\noindent
In this section we demonstrate the applicability of the proposed MISDP-formulation and use it to
assess the quality of the relaxations proposed in~\cite{Asp08} and~\cite{Asp07}.
For solving the mixed integer semidefinite programs we use the MISDP solver SCIP-SDP~\cite{SCIP-SDP},
originally developed in~\cite{Mar13} and~\cite{MS12}. It combines the
branch and bound framework SCIP 3.2.1 \cite{SCIP} with 
interior-point SDP-solvers, in our case SDPA~7.3.8~\cite{SDPA6,SDPA7}.

@@

The tests were performed on a Linux cluster with Intel i3 CPUs with 3.2GHz, 4MB cache, and 8GB memory running
Linux. Each computation was performed single-threaded with a single process running on each computer and with 
a time limit of four hours. The code was compiled with gcc 4.4.5 with \texttt{-O3} optimization. For 39 matrices and all 15 choices of $\rho$,
the evaluation of~\eqref{Asp08Dual} was aborted because of the memory limit of 8GB. These are counted with the maximum allowed time of four hours.

@@

The solving times in seconds for the left- and right-hand side of the RIP are given in
Table~\ref{lhsRhsTime}. The relaxation~\eqref{Asp07}
is very fast, taking only a few seconds for all instances. The mixed integer semidefinite program obviously
cannot compete with these solving times, although for band-matrices, the difference is relatively
small. The Lagrangian Relaxation~\eqref{Asp08}, however, takes more time to produce bounds than is needed
by~\eqref{MISDP} to compute exact solutions. This is mainly caused by the fact that~\eqref{Asp08Dual} involves
$n+1$ matrix variables of dimension $n\times n$ with corresponding SDP-constraints instead of a 
single one each in~\eqref{MISDP} and~\eqref{Asp07}.
Furthermore, the big difference is also caused by the instances running into the 
memory limit, but even without those, the algorithm still takes slightly longer than~\eqref{MISDP}.

@@

\noindent
This paper shows that exact values of restricted isometry constants can be
computed via mixed integer semidefinite programming. The computed bounds
significantly improve the bounds computed by the semidefinite
relaxations~\eqref{Asp07} and~\eqref{Asp08}
proposed in~\cite{Asp07} and \cite{Asp08}, respectively. In conclusion, the relaxation~
\eqref{Asp07} is fast to compute, but produces a gap of around 90\,\% for
$\alpha_k$ and 30\,\% for $\beta_k$. If successfully solved, 
\eqref{Asp08} provides better bounds for $\beta_k$ for most types of 
matrices, but its running time is too large
to be used in practice. Moreover, directly solving~\eqref{MISDP} is often
faster and produces the exact value. Consequently, 
\eqref{Asp08} should not be used in practice.

Still, the matrices currently handable by~\eqref{MISDP} are small. For the
future, the hope is that with the advancement of solving techniques for
mixed integer semidefinite programs, the sizes of the matrices can be
significantly increased. The components to be improved include cutting
planes, primal heuristics, and branching rules.
